Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2023-01-08 supplement.tex
edits to main text, formatting
edits to main text
Small revisions to 3.1.1

\section{Model specification and fitting}
\label{modelspec}

@@

In our preliminary models, the unconstrained inferred split times between East
African populations and Europeans was inferred to have occurred around $40$ka,
and the inferred split between East and West African populations occurred about
$10$ka earlier. These recent split dates are at odds with archaeological
evidence for an earlier Out-of-Africa (OOA) expansion
\citep[e.g.]{Fu2014-am,Hublin2020-nk,Hajdinjak2021-jo}, with the estimated date
for Neanderthal admixture into Eurasian populations
\citep[e.g.][]{Reilly2022-ym}, and with previous genetics-based dating of
splits between African and Eurasian populations
\citep[e.g.]{Gutenkunst2009-ff,Jouganous2017-pq,Kamm2020-vn}. The parameter
uncertainty of our inferred divergence times (which accounts for the
non-independence of sites) does not account for these differences: in the
high-likelihood models, the $95\%$ confidence intervals for the East
African-European split was roughly $38 \pm 5$ka, and for the East/West African
split was $45 \pm 10$ka, with the uncertainties reflecting 2 standard
deviations according to bootstrap of genomic regions. We therefore expect these
discrepancies to result from model misspecification in the details of the
divergence between the ancestors of OOA and East African populations.

Our models suppose a single split leading to the OOA event followed by constant
symmetric migrations between eastern African and OOA populations, which is an
oversimplification. Previous studies that have focused on reconstructing the
OOA divergence time and subsequent migration rates typically included more
Eurasian populations and multiple epochs that allow for differing continuous
migration rates and population sizes. Since a detailed model for the OOA (and
Back-to-Africa) event was not our main focus, we chose to keep this simplified
model, but fix the East African-Eurasian split at $50$ka
\citep{Fewlass2020-se,Hublin2020-nk}. This also allowed us to include
Neanderthal admixture into the OOA population at $45$ka.

@@

\begin{table}[ht]
\caption{
    \label{tab:supp-merger-with-stem-migration}
    \textbf{Best-fit parameters from the Merger-With-Stem-Migration model.}
    Inferred values are scaled to physical units assuming a generation time of
    29 years. This model gave a log-likelihood of -101,618.
    \aprcomment{if the slightly shift, also fix elsewhere in supp and main text}
}
\centering
\begin{tabular}[t]{rP{8cm}cc}
    \toprule
    Parameter & Description & Value & Std. err.\\
    \midrule
    $N_e$ & Ancestral effective population size & 9640 & 6410--11910 \\
    $N_{stem1}$ & Size of stem 1 lineage between Neanderthal split and stem 1E--stem 1S split & 100 & 100--851 \\
    $N_{stem2}$ & Size of stem 2 lineage & 24210 & 22130--26340 \\
    $N_{Nama_0}$ & Initial Nama size & 13440 & 11500--13560 \\
    $N_{Nama_F}$ & Final Nama size & 219 & 214--231 \\
    $N_{MSL_0}$ & Initial Mende size & 11380 & 9690--11290 \\
    $N_{MSL_F}$ & Final Mende size & 27760 & 26820--32280 \\
    $N_{EA}$ & Size of East African Branch & 9100 & 7970--8860 \\
    $N_{Gumuz_F}$ & Final Gumuz size & 3410 & 3270--3570 \\
    $N_{EP}$ & East African agriculturist size & 13510 & 12780--13500 \\
    $N_{GBR_0}$ & Initial British size & 958 & 928--1002 \\
    $N_{GBR_F}$ & Final British size & 12300 & 11850--12600 \\
    $N_{Neand}$ & Neanderthal size & 2520 & 2100--2730 \\
    $T_{stems}$ & Stem split time (years) & 1692000 & 695000--1785000 \\
    $T_{stem1}$ & Stem 1S--stem 1E split time (years) & 478900 & 275600--478300 \\
    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $0.98\times10^{-5}$ & $0.66\times10^{-5}$--$1.57\times10^{-5}$ \\
    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.48\times10^{-5}$ & $4.45\times10^{-5}$--$5.52\times10^{-5}$ \\
    $m_{MSL-EA}$ & Mende--East Africa migration rate & $19.7\times10^{-5}$ & $20.0\times10^{-5}$--$21.7\times10^{-5}$ \\
    $m_{EA-GBR}$ & East Africa--Europe migration rate & $3.82\times10^{-5}$ & $3.25\times10^{-5}$--$4.27\times10^{-5}$ \\
    $m_{EA-EA}$ & Intra-East Africa migration rate & $35.9\times10^{-5}$ & $33.7\times10^{-5}$--$40.9\times10^{-5}$ \\
    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.640 & 0.636--0.667 \\
    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2ka & 0.259 & 0.237--0.267 \\
    $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.158 & 0.154--0.167 \\
    $m_{stems}$ & Stem 1--stem 2 migration rate & $12.6\times10^{-5}$ & $3.95\times10^{-5}$--$12.3\times10^{-5}$ \\
    $T_{Nama}$ & Time of Nama merger event & 119500 & 101300--124600 \\
    $f_{stem 2 \rightarrow Nama}$ & Proportion of stem 2 ancestry making up initial Nama lineage ($1-f$ from stem 1S) & 0.698 & 0.694--0.758 \\
    $T_{EA}$ & Time of East Africa merger event & 96300 & 95400--111000 \\
    $f_{stem 2 \rightarrow EA}$ & Proportion of stem 2 ancestry making up initial East Africa lineage ($1-f$ from stem 1E) & 0.515 & 0.514--0.589 \\
    $T_{MSL}$ & Time of secondary admixture from stem 2 to Mende & 25120 & 22210--26300 \\
    $f_{stem 2 \rightarrow MSL}$ & Proportion of ancestry from secondary stem 2 admixture to Mende & 0.187 & 0.182--0.211 \\
    \bottomrule
\end{tabular}
\end{table}

>>>>

\section{Model specification and fitting}
\label{sec:modelspec}

@@

In our preliminary models, the unconstrained inferred split times between East
African populations and Europeans was inferred to have occurred around $40$ka,
and the inferred split between East and West African populations occurred about
$10$ka earlier. These recent split dates are at odds with paleoanthropological 
evidence for an earlier Out-of-Africa (OOA) expansion
\citep[e.g.][]{Fu2014-am,Hublin2020-nk,Hajdinjak2021-jo}, with the estimated date
for Neanderthal admixture into Eurasian populations
\citep[e.g.][]{Reilly2022-ym}, and with previous genetics-based dating of
splits between African and Eurasian populations
\citep[e.g.]{Gutenkunst2009-ff,Jouganous2017-pq,Kamm2020-vn}. The parameter
uncertainty of our inferred divergence times (which accounts for the
non-independence of sites) does not account for these differences: in the
high-likelihood models, the $95\%$ confidence intervals for the East
African-European split was roughly $38 \pm 5$ka, and for the East/West African
split was $45 \pm 10$ka, with the uncertainties reflecting 2 standard
deviations according to bootstrap of genomic regions. We therefore expect these
discrepancies to result from model misspecification in the details of the
divergence between the ancestors of OOA and East African populations.

Our models suppose a single split leading to the OOA event followed by constant
symmetric migrations between eastern African and OOA populations, which is an
oversimplification. Previous studies that have focused on reconstructing the
OOA divergence time and subsequent migration rates typically included more
Eurasian populations and multiple epochs that allow for differing continuous
migration rates and population sizes. Since a detailed model for the OOA (and
Back-to-Africa) event was not our main focus, we chose to keep this simplified
model, but fix the East African-Eurasian split at $50$ka to be older than 
archaeological dates for when the OOA reached Europe 
\citep{Fewlass2020-se,Hublin2020-nk}. This also allowed us to include 
Neanderthal admixture into the OOA population at $45$ka.

@@

\begin{table}[ht]
\caption{
    \label{tab:supp-merger-with-stem-migration}
    \textbf{Best-fit parameters from the Merger-With-Stem-Migration model.}
    Inferred values are scaled to physical units assuming a generation time of
    29 years. This model gave a log-likelihood of -101,618.
}
\centering
\begin{tabular}[t]{rP{8cm}cc}
    \toprule
    Parameter & Description & Value & Std. err.\\
    \midrule
    $N_e$ & Ancestral effective population size & 9640 & 6410--11910 \\
    $N_{stem1}$ & Size of stem 1 lineage between Neanderthal split and stem 1E--stem 1S split & 100 & 100--851 \\
    $N_{stem2}$ & Size of stem 2 lineage & 24210 & 22130--26340 \\
    $N_{Nama_0}$ & Initial Nama size & 13440 & 11500--13560 \\
    $N_{Nama_F}$ & Final Nama size & 219 & 214--231 \\
    $N_{MSL_0}$ & Initial Mende size & 11380 & 9690--11290 \\
    $N_{MSL_F}$ & Final Mende size & 27760 & 26820--32280 \\
    $N_{EA}$ & Size of East African Branch & 9100 & 7970--8860 \\
    $N_{Gumuz_F}$ & Final Gumuz size & 3410 & 3270--3570 \\
    $N_{EP}$ & East African agriculturist size & 13510 & 12780--13500 \\
    $N_{GBR_0}$ & Initial British size & 958 & 928--1002 \\
    $N_{GBR_F}$ & Final British size & 12300 & 11850--12600 \\
    $N_{Neand}$ & Neanderthal size & 2520 & 2100--2730 \\
    $T_{stems}$ & Stem split time (years) & 1692000 & 695000--1785000 \\
    $T_{stem1}$ & Stem 1S--stem 1E split time (years) & 478900 & 275600--478300 \\
    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $0.98\times10^{-5}$ & $0.66\times10^{-5}$--$1.57\times10^{-5}$ \\
    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.48\times10^{-5}$ & $4.45\times10^{-5}$--$5.52\times10^{-5}$ \\
    $m_{MSL-EA}$ & Mende--East Africa migration rate & $19.7\times10^{-5}$ & $20.0\times10^{-5}$--$21.7\times10^{-5}$ \\
    $m_{EA-GBR}$ & East Africa--Europe migration rate & $3.82\times10^{-5}$ & $3.25\times10^{-5}$--$4.27\times10^{-5}$ \\
    $m_{EA-EA}$ & Intra-East Africa migration rate & $35.9\times10^{-5}$ & $33.7\times10^{-5}$--$40.9\times10^{-5}$ \\
    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.640 & 0.636--0.667 \\
    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2ka & 0.259 & 0.237--0.267 \\
    $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.158 & 0.154--0.167 \\
    $m_{stems}$ & Stem 1--stem 2 migration rate & $12.6\times10^{-5}$ & $3.95\times10^{-5}$--$12.3\times10^{-5}$ \\
    $T_{Nama}$ & Time of Nama merger event & 119500 & 101300--124600 \\
    $f_{stem 2 \rightarrow Nama}$ & Proportion of stem 2 ancestry making up initial Nama lineage ($1-f$ from stem 1S) & 0.698 & 0.694--0.758 \\
    $T_{EA}$ & Time of East Africa merger event & 96300 & 95400--111000 \\
    $f_{stem 2 \rightarrow EA}$ & Proportion of stem 2 ancestry making up initial East Africa lineage ($1-f$ from stem 1E) & 0.515 & 0.514--0.589 \\
    $T_{MSL}$ & Time of secondary admixture from stem 2 to Mende & 25120 & 22210--26300 \\
    $f_{stem 2 \rightarrow MSL}$ & Proportion of ancestry from secondary stem 2 admixture to Mende & 0.187 & 0.182--0.211 \\
    \bottomrule
\end{tabular}
\end{table}
